EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-09-27 jlamp/jlamp.tex
minima

Let us start with the case of general hypergraphs. These were introduced in~\cite{concur2006}, even if no general consideration about their structure as a category was proved, and adhesivity, 
which is the main focus here, was yet to be presented to the world.

@@

\begin{defi}
	Let $\mathcal{G}$ be an hypergraph with equivalence, the category $\EqHyps$ of \emph{labelled hypergraph with equivalence}  is $\EqHyp/\mathcal{G}$.
\end{defi}

\begin{rem} The equivalence relation on the  nodes of the labelling hypergraph put constraints on the possible relations on the labelled hypergraphs. Indeed, consider the following hypergraphs with equivalences $\mathcal{G}$ and $\mathcal{H}$.

\noindent
\begin{minipage}[l]{.6\linewidth}
	\setlength{\parindent}{1.5em}
	Notice that there is no morphism $\mathcal{H}\to \mathcal{G}$. Indeed $\id{T(\mathcal{G})}$ is the unique morphism $T(\mathcal{G})\to T(\mathcal{H})$ while there are only two morphisms $f, g\colon Q_{\mathcal{H}}\rightrightarrows Q_{\mathcal{G}}$: the first one is constant in $v$, while the second one is constant in $w$.
\end{minipage}\hfill \qquad 
\begin{minipage}[r]{.4\linewidth}
\begin{tikzpicture}[baseline=(v1)]
	\begin{pgfonlayer}{nodelayer}
		\node[style=small box](a1) at (-.5, 1){};
		\node[style=none] (a1att) at (-.5, 1){};
		\node[style=node](v1) at (-.5, 0){};
		\node[style=small box](a2) at (.5, 1){};
		\node[style=none] (a2in) at (.22, 1){};
		\node[style=none] (a2out) at (.78, 1){};
		\node[style=node](v2) at (.5, 0){};
		\node[style=none] (l1) at (-.5, 1.5){$a$};
		\node[style=none](l2) at (.5, 1.5){$b$};
		\node[style=none] (l1) at (-1, 0){$v$};
		\node[style=none](l2) at (1, 0){$w$};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw(a1att.south) to (v1);
		\draw[style=none, out=180, in=180](v2) to (a2in.west);
		\draw[style=none, out=0, in=0](v2) to (a2out.east);
	\end{pgfonlayer}
	\begin{pgfonlayer}{eqlayer}
		\draw[dashed, rounded corners] (-.8, -0.3) rectangle (-.2, 0.3);
		\draw[dashed, rounded corners] (.2, -0.3) rectangle (.8, 0.3);
	\end{pgfonlayer}
\end{tikzpicture}\qquad  
\begin{tikzpicture}[baseline=(v1)]
	\begin{pgfonlayer}{nodelayer}
		\node[style=small box](a1) at (-.5, 1){};
		\node[style=none] (a1att) at (-.5, 1){};
		\node[style=node](v1) at (-.5, 0){};
		\node[style=small box](a2) at (.5, 1){};
		\node[style=none] (a2in) at (.22, 1){};
		\node[style=none] (a2out) at (.78, 1){};
		\node[style=node](v2) at (.5, 0){};
		\node[style=none] (l1) at (-.5, 1.5){$a$};
		\node[style=none](l2) at (.5, 1.5){$b$};
		\node[style=none] (l1) at (-1, 0){$v$};
		\node[style=none](l2) at (1, 0){$w$};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw(a1att.south) to (v1);
		\draw[style=none, out=180, in=180](v2) to (a2in.west);
		\draw[style=none, out=0, in=0](v2) to (a2out.east);
	\end{pgfonlayer}
	\begin{pgfonlayer}{eqlayer}
		\draw[dashed, rounded corners] (-.8, -0.3) rectangle (.8, 0.3);
	\end{pgfonlayer}
\end{tikzpicture}
\end{minipage}

 For both cases we have:
\[ f(q_{\mathcal{H}}(w))=v \neq w =q_{\mathcal{G}}(w)=q_\mathcal{G}(\id{V_{\mathcal{G}}}(w)) \qquad  g(q_{\mathcal{H}}(v))=w \neq v =q_{\mathcal{G}}(v)=q_\mathcal{G}(\id{V_{\mathcal{G}}}(v))  \]

	Intuitively, labelling with $\mathcal{G}$ force us to keep all the sorts (i.e.~the labellings of the nodes) distinguished, while labelling with $\mathcal{H}$ allows us to merge different sorts.
\end{rem}

@@

\begin{defi}Let $\mathcal{G}$ be an hypergraph and $(\mathcal{H},(f,g))$ a labelled hypergraph with equivalence.  We say that $(\mathcal{H},(f,g))$ is a \emph{left-monogamous} (resp.~\emph{separated, pruned}) $\mathcal{G}$-hypergraph if $\mathcal{H}$ is left-monogamous (resp.~separated, pruned).

@@

 In particular, the previous  proposition, together with Lemmas \ref{lem:equ}, \ref{lem:coequ}, \ref{colim} and \Cref{lim}, yields the following

@@

\begin{restatable}{prop}{eh}\label{prop:eh}
Consider an hypergraph with equivalence $\mathcal{G} = (E, V, Q, s, t, q)$ and let $(A, \pi^A_1, \pi^A_2)$, $(B, \pi^B_1, \pi^B_2)$ be kernel pairs for, respectively, $q^\star\circ s$ and $q^\star\circ t$. Then the following are equivalent:
\begin{enumerate}
	\item $\mathcal{G}$ is an e-hypergraph;
	\item $q^\star\circ t\circ \pi^A_1=q^\star\circ t\circ \pi^A_2$;
	\item there exists a unique arrow $k\colon A\to B$ such that 
	\[\pi^B_1\circ k = \pi^A_1 \qquad \pi^B_2\circ k = \pi^A_2 \]
	\end{enumerate}\end{restatable}

@@

Let now $\pbe$ be the class of arrows in $\egg$ that are sent to $\pbc$ by $E$. By definition and \Cref{lem:pbmono}, we have that such class is closed under composition and decomposition, it contains all isomorphisms and it is stable under pullbacks. Using \Cref{lem:pbmono} and \Cref{lem:epo} we can further deduce that it is stable under pushouts. Then \Cref{thm:slice-functors} and \Cref{cor:eqade} yield the following.

@@

\begin{defi}We will define $\elm$, $\eps$ as the full subcategories of $\egg$ given by those e-hypergraphs which are, respectively, left-monogamous and pruned and separated. We will denote by $M_{\mathsf{e}}$ and by $S_{\mathsf{e}}$ the corresponding functors.
\end{defi}

@@

We begin by providing an equivalent characterization to labelled e-hypergraphs (see \Cref{proof:fh} for details).

\begin{restatable}{lem}{fh}\label{lem:fh}
Let $(a,b,c)\colon \mathcal{H}\to \mathcal{G}$ be an object of $\hyp_{\eq, \mathcal{G}}$ and let $(A, \pi^A_1, \pi^A_2)$, $(B, \pi^B_1, \pi^B_2)$ be kernel pairs for, respectively, $(q^\star\circ s_{\mathcal{H}}, a),(q^\star\circ t_{\mathcal{H}}, a)\colon E_{\mathcal{H}}\to Q_{\mathcal{H}}\times E_{\mathcal{G}}$. Then the following are equivalent:
\begin{enumerate}
	\item $(a,b,c)$ is a labelled e-hypergraph;
	\item $q^\star\circ t_{\mathcal{H}}\circ \pi^A_1=q^\star\circ t_{\mathcal{H}}\circ \pi^A_2$;
	\item there exists a unique arrow $k\colon A\to B$ such that 
	\[\pi^B_1\circ k = \pi^A_1 \qquad \pi^B_2\circ k = \pi^A_2 \]
\end{enumerate}
\end{restatable}

@@

\begin{restatable}{lem}{varie}\label{prop:varie}
	Let $\mathcal{G}$ be an hypergraph with equivalence, then $\eggl$ is closed in $\hyp_{\eq, \mathcal{G}}$ under connected limits and pushouts alongs arrows in $\pbe_\mathcal{G}$, where $\pbe_\mathcal{G}$ is the class of morphisms in $\eggl$ whose image through $E_\mathcal{G}$ is in $\pbc_\mathcal{G}$. 
\end{restatable}

@@

Putting together \Cref{cor:eti} with \Cref{prop:varie} we get the following.

@@

The fact that the right-hand side is a regular mono ensures that if the e-hypergraph to whom a rule is applied is acyclic,
the same property holds for the e-hypergraph obtained as the result.
%
Note however that cycles may implicitly occur due to the presence of the equivalence 
on nodes for what are called collapsing rules, i.e. where the right-hand side is a single variable. 
%
Looking again at the key example of~\cite{WillseyNWFTP21}, consider the rule $1 \times x = x$,
which can be modelled as the DPO rule below
\[
\hspace{.25cm}
\xymatrix{        
	\begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v1) at (-0.9, 0.4){};
			\node[style=node](v2) at (-0.9, -0.4){};
			\node[style=medium box] at (0, 0){$\times$};
			\node[style=none](divfst) at (-0.3, 0.3){};
			\node[style=none](divsnd) at (-0.3, -0.3){};
			\node[style=none](divout) at (0.3, 0){};
			\node[style=node] (w) at (0.9, 0){};
			\node[style=small box] at (-2.3, 0.3){$1$};
			\node[style=none](one) at (-2, 0.3){};
			\node[style=node](z) at (-1.4, 0.4){};
		\end{pgfonlayer}        
		\begin{pgfonlayer}{edgelayer}
			\draw(v1) to (divfst.center);
			\draw(v2) to (divsnd.center);
			\draw(divout.center) to (w);
			\draw(one.center) to (z);
		\end{pgfonlayer}
		\begin{pgfonlayer}{eqlayer}
			\draw[dashed, rounded corners](-1.7, 0.1) rectangle (-0.6, 0.7);
			\draw[dashed, rounded corners](1.2, -0.3) rectangle (0.6, 0.3);
			\draw[dashed, rounded corners] (-1.2, -0.7) rectangle (-.6, -0.1);
		\end{pgfonlayer}\begin{pgfonlayer}{background}
			\draw[color=white] (-1.5, 0.1) rectangle (1.5, -0.3);
		\end{pgfonlayer}
	\end{tikzpicture}
	\ar@{=>}[r] &	\hspace{12pt}
	\begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v1) at (-0.9, 0.3){};
			\node[style=node](v2) at (-0.9, -0.3){};
			\node[style=medium box] at (0, 0){$\times$};
			\node[style=none](divfst) at (-0.3, 0.3){};
			\node[style=none](divsnd) at (-0.3, -0.3){};
			\node[style=none](divout) at (0.3, 0){};
			\node[style=node] (w) at (0.9, 0){};
			\node[style=small box] at (-2.3, 0.3){$1$};
			\node[style=none](one) at (-2, 0.3){};
			\node[style=node](z) at (-1.4, 0.3){};
		\end{pgfonlayer}        
		\begin{pgfonlayer}{edgelayer}
			\draw(v1) to (divfst.center);
			\draw(v2) to (divsnd.center);
			\draw(divout.center) to (w);
			\draw(one.center) to (z);
		\end{pgfonlayer}
		\begin{pgfonlayer}{eqlayer}
			\draw[dashed, rounded corners](-1.7, 0.6) rectangle (-0.6, -0);
		\end{pgfonlayer}
		\begin{pgfonlayer}{background}
			\draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
		\end{pgfonlayer}
		\draw[dashed,rounded corners]
		([shift={(0,0.2)}]v2.north) --
		([shift={(-0.2,0)}]v2.west) --
		(-.9,-1.2)--(.9,-1.2)--([shift={(0.2,0)}]w.east)--
		([shift={(0,0.2)}]w.north)  -- ([shift={(-0.2,0)}]w.west)-- (0.7,-0.8)--(-0.7,-0.8)--([shift={(0.2,0)}]v2.east)-- cycle;
	\end{tikzpicture}
}
\]
%
Intuitively, the e-term graph obtained after the application of the rule seem to represent a recursive term such 
as $rec_x. 1 \times x$, highlighting the difficulty of extracting a term representation out of an e-term graph.

The fact that the left-hand side of a rule is the identity allows for the repeated application of such rule, hence the 
possibility to keep on performing the same rewriting step.
This is forbidden by using rules \emph{with negative application conditions}, given as the usual span 
$L \leftarrow K \rightarrow R$ plus an additional arrow $n: L\rightarrow N$, such that a match $m: L \to G$ is admissible if it cannot
be factorised through $n$~\cite{HabelHT96}. For e-term graphs and rules $L \leftarrow L \rightarrow R$, it suffices to choose $n$ 
as the right-hand side itself.
Indeed, consider again the rule $x / x \to 1$: once $1$ has been added once to the equivalence class of $x / x$, it is not necessary to
keep on adding it, since e.g. the two pictures below intuitively represent the same EGG
\[xxx\]
%
The theory of $\mathcal{M}$-adhesivity carries on for rules with negative application conditions, see~\cite{ehrig2012,ehrig2014adhesive}.

@@

\emph{Related works.}
%\todo{espandere un poco}
Despite the interest they have ben raising as an efficient data structure, we are not aware of any attempt to provide an 
algebraic characterisation of EGGs, the only exception being~\cite{ghica}. We leave for future work an in-depth comparison 
among the two proposals, which appear to be related yet quite different.
%
In fact, for both proposals the key intuition is  to use string diagrams to represents term graphs. We adopted a
more down-to-earth approach by equipping
concretely the nodes of a term graph with an equivalence relation, thus directly corresponding to the original presentation~\cite{DetlefsNS05},
at the same time extending and generalising it~\cite{concur2006} in the contemporary jargon of adhesive categories. 
The route chosen in~\cite{ghica} is to consider term graphs as arrows of a symmetric monoidal category,
and equipping them with an enrichment over semi-lattices, the resulting formalism being reminiscent of hierarchical hypergraphs.
Thus, the solution in~\cite{ghica}  is  more general than ours since it can be lifted to term graphs defined over categories other than
$\Set$. At the same time, we both consider the DPO approach for rewriting, even if only our solution guarantees that the resulting category
is actually $\mathcal{M}$-adhesive, thus allowing to exploit all the features of the framework
as they hold for DPO graph transformation.  Moreover, our construction is technically simpler, which facilitates its practical application 
and theoretical analysis, and aligns with the traditional methodology adopted in the graph transformation community 
of explicitly endowing (hyper)graphs with additional structure.
%\section*{Acknowledgments}
%Hypergraphs, terms graphs, e-graphs, adhesive categories.

>>>>

Let us start with the case of general hypergraphs. These were introduced in~\cite{concur2006}, even if no general consideration about their structure as a category was provided, and adhesivity, 
which is the main focus here, was yet to be presented to the world.

@@

\begin{defi}
	Let $\mathcal{G}$ be an hypergraph with equivalence. The category $\EqHyps$ of hypergraph with equivalence \emph{labelled by $\mathcal{G}$}  is $\EqHyp/\mathcal{G}$.
\end{defi}

\begin{rem} The equivalence relation on the  nodes of the labelling hypergraph put constraints on the relations between the labelled hypergraphs. Consider the following hypergraphs with equivalences $\mathcal{G}$ and $\mathcal{H}$.

\smallskip
\noindent
\begin{minipage}[l]{.6\linewidth}
	\setlength{\parindent}{1.5em}
	Notice that there is no morphism $\mathcal{H}\to \mathcal{G}$. Indeed $\id{T(\mathcal{G})}$ is the unique morphism $T(\mathcal{G})\to T(\mathcal{H})$ while there are only two morphisms $f, g\colon Q_{\mathcal{H}}\rightrightarrows Q_{\mathcal{G}}$: the first one is constant in $v$, while the second one is constant in $w$.
\end{minipage}\hfill \qquad 
\begin{minipage}[r]{.4\linewidth}
\begin{tikzpicture}[baseline=(v1)]
	\begin{pgfonlayer}{nodelayer}
		\node[style=small box](a1) at (-.5, 1){};
		\node[style=none] (a1att) at (-.5, 1){};
		\node[style=node](v1) at (-.5, 0){};
		\node[style=small box](a2) at (.5, 1){};
		\node[style=none] (a2in) at (.22, 1){};
		\node[style=none] (a2out) at (.78, 1){};
		\node[style=node](v2) at (.5, 0){};
		\node[style=none] (l1) at (-.5, 1.5){$a$};
		\node[style=none](l2) at (.5, 1.5){$b$};
		\node[style=none] (l1) at (-1, 0){$v$};
		\node[style=none](l2) at (1, 0){$w$};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw(a1att.south) to (v1);
		\draw[style=none, out=180, in=180](v2) to (a2in.west);
		\draw[style=none, out=0, in=0](v2) to (a2out.east);
	\end{pgfonlayer}
	\begin{pgfonlayer}{eqlayer}
		\draw[dashed, rounded corners] (-.8, -0.3) rectangle (-.2, 0.3);
		\draw[dashed, rounded corners] (.2, -0.3) rectangle (.8, 0.3);
	\end{pgfonlayer}
\end{tikzpicture}\qquad  
\begin{tikzpicture}[baseline=(v1)]
	\begin{pgfonlayer}{nodelayer}
		\node[style=small box](a1) at (-.5, 1){};
		\node[style=none] (a1att) at (-.5, 1){};
		\node[style=node](v1) at (-.5, 0){};
		\node[style=small box](a2) at (.5, 1){};
		\node[style=none] (a2in) at (.22, 1){};
		\node[style=none] (a2out) at (.78, 1){};
		\node[style=node](v2) at (.5, 0){};
		\node[style=none] (l1) at (-.5, 1.5){$a$};
		\node[style=none](l2) at (.5, 1.5){$b$};
		\node[style=none] (l1) at (-1, 0){$v$};
		\node[style=none](l2) at (1, 0){$w$};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw(a1att.south) to (v1);
		\draw[style=none, out=180, in=180](v2) to (a2in.west);
		\draw[style=none, out=0, in=0](v2) to (a2out.east);
	\end{pgfonlayer}
	\begin{pgfonlayer}{eqlayer}
		\draw[dashed, rounded corners] (-.8, -0.3) rectangle (.8, 0.3);
	\end{pgfonlayer}
\end{tikzpicture}
\end{minipage}

 For both cases we have
\[ f(q_{\mathcal{H}}(w))=v \neq w =q_{\mathcal{G}}(w)=q_\mathcal{G}(\id{V_{\mathcal{G}}}(w)) \qquad  g(q_{\mathcal{H}}(v))=w \neq v =q_{\mathcal{G}}(v)=q_\mathcal{G}(\id{V_{\mathcal{G}}}(v))  \]

	Intuitively, labelling with $\mathcal{G}$ force us to keep all the sorts (i.e.~the labelling of the nodes) distinguished, while labelling with $\mathcal{H}$ allows us to merge different sorts.
\end{rem}

@@

\begin{defi}Let $\mathcal{G}$ be an hypergraph and $(\mathcal{H},(f,g))$ a labelled hypergraph with equivalence.  We say that $(\mathcal{H},(f,g))$ is a \emph{left-monogamous} (resp.~\emph{separated, pruned}) $\mathcal{G}$-hypergraph with equivalence if $\mathcal{H}$ is left-monogamous (resp.~separated, pruned).

@@

 In particular, the previous proposition, together with Lemmas \ref{lem:equ}, \ref{lem:coequ}, \ref{colim} and \Cref{lim}, yields the following result.

@@

\begin{restatable}{prop}{eh}\label{prop:eh}
Let $\mathcal{G} = (E, V, Q, s, t, q)$ be an hypergraph with equivalence and $(A, \pi^A_1, \pi^A_2)$, $(B, \pi^B_1, \pi^B_2)$ kernel pairs for, respectively, $q^\star\circ s$ and $q^\star\circ t$. Then the following facts are equivalent
\begin{enumerate}
	\item $\mathcal{G}$ is an e-hypergraph;
	\item $q^\star\circ t\circ \pi^A_1=q^\star\circ t\circ \pi^A_2$;
	\item there exists a unique arrow $k\colon A\to B$ such that 
	\[\pi^B_1\circ k = \pi^A_1 \qquad \pi^B_2\circ k = \pi^A_2 \]
	\end{enumerate}\end{restatable}

@@

Let now $\pbe$ be the class of arrows in $\egg$ that are sent to $\pbc$ by $E$. By definition and \Cref{lem:pbmono}, we have that such class is closed under composition and decomposition, it contains all isomorphisms and it is stable under pullbacks. Using \Cref{lem:pbmono} and \Cref{lem:epo} we can further deduce that it is stable under pushouts. Then \Cref{thm:slice-functors} and \Cref{cor:eqade} yield the following result.

@@

\begin{defi}We will define $\elm$ and $\eps$ as the full subcategories of $\egg$ given by those e-hypergraphs that are, respectively, left-monogamous and pruned and separated. We will denote by $M_{\mathsf{e}}$ and by $S_{\mathsf{e}}$ the corresponding functors.
\end{defi}

@@

\smallskip
We now provide an equivalent characterisation to labelled e-hypergraphs (see \Cref{proof:fh} for details).

\begin{restatable}{lem}{fh}\label{lem:fh}
Let $(a,b,c)\colon \mathcal{H}\to \mathcal{G}$ be an object of $\hyp_{\eq, \mathcal{G}}$ and $(A, \pi^A_1, \pi^A_2)$, $(B, \pi^B_1, \pi^B_2)$ kernel pairs for, respectively, $(q^\star\circ s_{\mathcal{H}}, a),(q^\star\circ t_{\mathcal{H}}, a)\colon E_{\mathcal{H}}\to Q_{\mathcal{H}}\times E_{\mathcal{G}}$. Then the following are equivalent
\begin{enumerate}
	\item $(a,b,c)$ is a labelled e-hypergraph;
	\item $q^\star\circ t_{\mathcal{H}}\circ \pi^A_1=q^\star\circ t_{\mathcal{H}}\circ \pi^A_2$;
	\item there exists a unique arrow $k\colon A\to B$ such that 
	\[\pi^B_1\circ k = \pi^A_1 \qquad \pi^B_2\circ k = \pi^A_2 \]
\end{enumerate}
\end{restatable}

@@

\begin{restatable}{lem}{varie}\label{prop:varie}
	Let $\mathcal{G}$ be an hypergraph with equivalence. Then $\eggl$ is closed in $\hyp_{\eq, \mathcal{G}}$ under connected limits and pushouts alongs arrows in $\pbe_\mathcal{G}$, where $\pbe_\mathcal{G}$ is the class of morphisms in $\eggl$ whose image through $E_\mathcal{G}$ is in $\pbc_\mathcal{G}$. 
\end{restatable}

@@

Putting together \Cref{cor:eti} with \Cref{prop:varie} we get the following result.

@@

The fact that the right-hand side is a regular mono ensures that if the e-hypergraph to whom a rule is applied is acyclic,
the same property holds for the e-hypergraph obtained as the result.
%
Note however that cycles may implicitly occur due to the presence of the equivalence 
on nodes for what are called collapsing rules, i.e. where the right-hand side is a single variable. 
%
Looking again at the key example of~\cite{WillseyNWFTP21}, consider the rule $1 \times x = x$,
which can be modelled as the DPO rule below
\[
\hspace{.25cm}
\xymatrix{        
	\begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v1) at (-0.9, 0.4){};
			\node[style=node](v2) at (-0.9, -0.4){};
			\node[style=medium box] at (0, 0){$\times$};
			\node[style=none](divfst) at (-0.3, 0.3){};
			\node[style=none](divsnd) at (-0.3, -0.3){};
			\node[style=none](divout) at (0.3, 0){};
			\node[style=node] (w) at (0.9, 0){};
			\node[style=small box] at (-2.3, 0.3){$1$};
			\node[style=none](one) at (-2, 0.3){};
			\node[style=node](z) at (-1.4, 0.4){};
		\end{pgfonlayer}        
		\begin{pgfonlayer}{edgelayer}
			\draw(v1) to (divfst.center);
			\draw(v2) to (divsnd.center);
			\draw(divout.center) to (w);
			\draw(one.center) to (z);
		\end{pgfonlayer}
		\begin{pgfonlayer}{eqlayer}
			\draw[dashed, rounded corners](-1.7, 0.1) rectangle (-0.6, 0.7);
			\draw[dashed, rounded corners](1.2, -0.3) rectangle (0.6, 0.3);
			\draw[dashed, rounded corners] (-1.2, -0.7) rectangle (-.6, -0.1);
		\end{pgfonlayer}\begin{pgfonlayer}{background}
			\draw[color=white] (-1.5, 0.1) rectangle (1.5, -0.3);
		\end{pgfonlayer}
	\end{tikzpicture}
	\ar@{=>}[r] &	\hspace{12pt}
	\begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v1) at (-0.9, 0.3){};
			\node[style=node](v2) at (-0.9, -0.3){};
			\node[style=medium box] at (0, 0){$\times$};
			\node[style=none](divfst) at (-0.3, 0.3){};
			\node[style=none](divsnd) at (-0.3, -0.3){};
			\node[style=none](divout) at (0.3, 0){};
			\node[style=node] (w) at (0.9, 0){};
			\node[style=small box] at (-2.3, 0.3){$1$};
			\node[style=none](one) at (-2, 0.3){};
			\node[style=node](z) at (-1.4, 0.3){};
		\end{pgfonlayer}        
		\begin{pgfonlayer}{edgelayer}
			\draw(v1) to (divfst.center);
			\draw(v2) to (divsnd.center);
			\draw(divout.center) to (w);
			\draw(one.center) to (z);
		\end{pgfonlayer}
		\begin{pgfonlayer}{eqlayer}
			\draw[dashed, rounded corners](-1.7, 0.6) rectangle (-0.6, -0);
		\end{pgfonlayer}
		\begin{pgfonlayer}{background}
			\draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
		\end{pgfonlayer}
		\draw[dashed,rounded corners]
		([shift={(0,0.2)}]v2.north) --
		([shift={(-0.2,0)}]v2.west) --
		(-.9,-1.2)--(.9,-1.2)--([shift={(0.2,0)}]w.east)--
		([shift={(0,0.2)}]w.north)  -- ([shift={(-0.2,0)}]w.west)-- (0.7,-0.8)--(-0.7,-0.8)--([shift={(0.2,0)}]v2.east)-- cycle;
	\end{tikzpicture}
}
\]
%
Intuitively, the e-term graph obtained after the application of the rule seems to represent a recursive term such 
as $rec_x. 1 \times x$, highlighting the difficulty of extracting a term representation out of an e-term graph.

The fact that the left-hand side of a rule is the identity allows for the repeated application of such rule, hence the 
possibility to keep on performing the same rewriting step.
This is forbidden by using rules \emph{with negative application conditions}, given as the usual span 
$L \leftarrow K \rightarrow R$ plus an additional arrow $n: L\rightarrow N$, such that a match $m: L \to G$ is admissible if it cannot
be factorised through $n$~\cite{HabelHT96}. For e-term graphs and rules $L \leftarrow L \rightarrow R$, it suffices to choose $n$ 
as the right-hand side itself.
If you consider e.g. the rule $x / x \to 1$, once $1$ has been added once to the equivalence class of $x / x$, it is not necessary to
keep on adding it.
% since e.g. the two pictures below intuitively represent the same EGG
%\[xxx\]
%
The theory of $\mathcal{M}$-adhesivity carries on for rules with negative application conditions, see~\cite{ehrig2012,ehrig2014adhesive}.

@@

\emph{Related works.}
%\todo{espandere un poco}
Despite the interest they have ben raising as an efficient data structure, we are not aware of any attempt to provide an 
algebraic characterisation of EGGs, the only exception being~\cite{ghica}. We leave for future work an in-depth comparison 
among the two proposals, which appear to be related yet quite different.
%
%In fact, for both proposals the key intuition is  to use string diagrams to represents term graphs. 
We adopted a more down-to-earth approach by equipping
concretely the nodes of a term graph with an equivalence relation, thus directly corresponding to the original presentation~\cite{DetlefsNS05},
at the same time extending and generalising it~\cite{concur2006} in the contemporary jargon of adhesive categories. 
The route chosen in~\cite{ghica} is to consider term graphs as arrows of a symmetric monoidal category,
and equipping them with an enrichment over semi-lattices, the resulting formalism being reminiscent of hierarchical hypergraphs.
Thus, the solution in~\cite{ghica}  is  more general than ours since it can be lifted to term graphs defined over categories other than
$\Set$. At the same time, we both consider the DPO approach for rewriting, even if only our solution guarantees that the resulting category
is $\mathcal{M}$-adhesive, thus allowing to exploit the features of the framework
as they hold for DPO graph transformation.  Moreover, our construction is technically simpler, which facilitates its practical application 
and theoretical analysis, and aligns with the traditional methodology adopted in the graph transformation community 
of explicitly endowing (hyper)graphs with additional structure.
%\section*{Acknowledgments}
%Hypergraphs, terms graphs, e-graphs, adhesive categories.
